---
title: "NOAA Storm Data Analysis"
author: "cha"
date: "05/06/2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

#Synopsis
This analysis identifies the most harmful types of events for population health in the US (meaning fatalities and injuries)
 as well as the events with the greatest economic consequences (in USD). The analysis is based on the NOAA storm database. 

1. Most harmful types of events for population health in the US 
2. Events with the greatest economic consequences across the US

The documentation of database can be accessed here:https://d396qusza40orc.cloudfront.net/repdata%2Fpeer2_doc%2Fpd01016005curr.pdf

#Data Processing
    ```{r}
#load packages for data analysis
library(magrittr)
library(dplyr)
library(ggplot2)
library(tidyverse)
#read data
data <- read.csv("repdata_data_StormData.csv.bz2")
#summary(data$EVTYPE)
#table(data$EVTYPE)
#inspect data:
head(data)
#table(data$EVTYPE,data$INJURIES)
#Note: Fatalities and injuries are split into two columns for direct and indirect fatality/injury

#clean data
#combine event that are listed multiple times, different spelling
data$EVTYPE <- as.character(data$EVTYPE)
data$EVTYPE[grepl("/|&|and", data$EVTYPE,ignore.case = TRUE)] <- "Multiple Event"
data$EVTYPE[grepl("volc", data$EVTYPE,ignore.case = TRUE)] <- "Volcano"
data$EVTYPE[grepl("wind|wnd", data$EVTYPE,ignore.case = TRUE)] <- "WIND"
data$EVTYPE[grepl("FLOOD", data$EVTYPE,ignore.case = TRUE)] <- "FLOOD"
data$EVTYPE[grepl("storm", data$EVTYPE,ignore.case = TRUE)] <- "STORM"
data$EVTYPE[grepl("hail", data$EVTYPE,ignore.case = TRUE)] <- "HAIL"
data$EVTYPE[grepl("fire", data$EVTYPE,ignore.case = TRUE)] <- "FIRE"
data$EVTYPE[grepl("heat", data$EVTYPE,ignore.case = TRUE)] <- "HEAT"
data$EVTYPE[grepl("tornado", data$EVTYPE,ignore.case = TRUE)] <- "TORNADO"
data$EVTYPE[grepl("drought", data$EVTYPE,ignore.case = TRUE)] <- "DROUGHT"


#Summarise required data into one variable
#8 - eventtype
#23 - fatalities
#24 - injuries
healthdata <- data[,(c(8,23:24))]
head(healthdata)

h<-healthdata %>% 
  group_by(EVTYPE) %>%
  summarize(sum_fatalities = sum(FATALITIES), 
            sum_injuries = sum(INJURIES))
#sort by sum_fatalities 
h_ar <- arrange(h,sum_fatalities)
tail(h_ar)
#Result is Tornado

#sort by sum_injuries 
h_inj <- arrange(h,sum_injuries)
tail(h_inj)
#Result is Tornados 

#arrange data
a<-arrange(h_inj, desc(sum_fatalities), by_group = TRUE)
#find the 10 highest values first for damage crops
#eco <- aggregate(sum_damage ~ e$EVTYPE, e, sum) [order(e$sum_damage),]
h_inju<-a[1:20,]
h_inju

#Plot
plot1 <- ggplot(data=h_inju, aes(x=EVTYPE))+
  geom_boxplot(aes(y=sum_fatalities), colour="red")+
  geom_boxplot(aes(y=sum_injuries), colour="blue")+
  theme(axis.text.x = element_text(angle = 90, hjust = 1))


#Most harmful types of events for population health in the US (EVTYPE variable)
#Answer is Tornados 

###PART2
#2. Events with the greatest economic consequences across the US
#For this question, "damage" needs to be assesed. 
#Propdmg - property damage 
#Alphabetical characters used to signify magnitude include “K” for thousands, “M” for millions, and “B” for billions

#PROPERTY DAMAGE
head(data$PROPDMG)
tail(data$PROPDMGEXP)
#K=1000, M= 1000000, B=1000000000
data$PROPDMGEXP[data$PROPDMGEXP=="K"]<- 1000
data$PROPDMGEXP[data$PROPDMGEXP=="M"]<- 1000000
data$PROPDMGEXP[data$PROPDMGEXP=="m"]<- 1000000
data$PROPDMGEXP[data$PROPDMGEXP=="B"]<- 1000000000
#change from character to numeric
data$PROPDMGEXP <- as.numeric(data$PROPDMGEXP)
#multiply prop damage times prop damage xp (K,M,B)
data$damage<-data$PROPDMG*data$PROPDMGEXP

#CROP DAMAGE
#K=1000, M= 1000000, B=1000000000
data$CROPDMGEXP[data$CROPDMGEXP=="K"]<- 1000
data$CROPDMGEXP[data$CROPDMGEXP=="M"]<- 1000000
data$CROPDMGEXP[data$CROPDMGEXP=="m"]<- 1000000
data$CROPDMGEXP[data$CROPDMGEXP=="B"]<- 1000000000
#change from character to numeric
data$CROPDMGEXP <- as.numeric(data$CROPDMGEXP)
#multiply prop damage times prop damage xp (K,M,B)
data$damagecrop<-data$CROPDMG*data$CROPDMGEXP

#select eventtype variable and cost of damage variable for property and crops 
economicdata <- data[,(c(8,38,39))]

#summarise 
xx<-economicdata %>% 
 group_by(EVTYPE) %>%  
  summarise(damagecrop, damage, na.rm=TRUE)
xx

#variable for property damage
#arrange data by max value for damage property 
x<-arrange(xx, desc(damage), by_group = FALSE)
#find the 10 highest values first for damage property
#eco <- aggregate(sum_damage ~ e$EVTYPE, e, sum) [order(e$sum_damage),]
ecop<-x[1:20,]
ecop

#variable for crop damage
#arrange data by max value for damage crops
y<-arrange(xx, desc(damagecrop), by_group = TRUE)
#find the 10 highest values first for damage crops
#eco <- aggregate(sum_damage ~ e$EVTYPE, e, sum) [order(e$sum_damage),]
ecoc<-y[1:20,]
ecoc

#Plot2 damage property by event type 
plot2 <- ggplot(data=ecop, aes(x=EVTYPE, y=(damage/1000000000)))+
  geom_line(colour="blue")+
  theme(axis.text.x = element_text(angle = 90, hjust = 1))

#Plot3 damage property by event type 
plot3 <- ggplot(data=ecoc, aes(x=EVTYPE, y=(damagecrop/1000000000)))+
  geom_line(colour="red")+
  theme(axis.text.x = element_text(angle = 90, hjust = 1))

```

#Results
1. The most harmful types of events for population health in the US are Tornadoes, as shown below. Red being fatalities, blue injuries.
```{r}
plot1
```


2. The events with the greatest economic consequences across the US are Floods (considering property) and Droughts considering (crops):
```{r}
plot2
ecoc
```

